Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-12-03 manuscript/manuscript.Rmd
starting to work on #7
closes #7 : creates table of chl a values across range of probability of exceeding and different advisory levels.

################################################################################
# Condtional Probability
################################################################################
mc_guides <- c(0.3, 1, 1.6, 2, 4, 10, 20)
mc_names <- c("USEPA","WHO","USEPA","WHO","WHO","WHO","WHO")
mc_type <- c("Child Drinking Water Advisory",
             "Drinking Water Advisory",
             "Adult Drinking Water",
             "Recreational: Low Prob. of Effect",
             "Recreational: Moderate Prob. of Effect",
             "Recreational: High Prob. of Effect",
             "Recreational: Very High Prob. of Effect")
mc_chla <- vector("numeric",length(mc_guides))
#wgt <- nla_dat$WGT_NLA
wgt<- rep(1,nrow(nla_dat))

@@

mc_chla[1] <- get_chla(epa_child_cp,0.5)

@@

mc_chla[2] <- get_chla(who_drink_cp,0.5)

@@

mc_chla[3] <- get_chla(epa_adult_cp,0.5)

@@

mc_chla[4] <- get_chla(who_rec_low1_cp,0.5)

@@

```{r mc_chla_table, results='asis', echo=FALSE, tab.cap="Chlorophyll \\textit{a} concentrations that are associated with a 50% probability of exceeding a microcystin-LR health advisory concentration. \\label{tab:mc_chla_table}"}
mc_chla_tbl <- data.frame(Source=mc_names,Type=mc_type,Microcystin=mc_guides,Chlorophyll=mc_chla)[1:4,]
#pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
knitr::kable(mc_chla_tbl)
####AS OF 12/2/15 - REDO THIS WITH MULTIPLE COND PROBS!
```

>>>>

################################################################################
# Condtional Probability
################################################################################
mc_guides <- c(0.3, 1, 1.6, 2, 4, 10, 20)
mc_names <- c("USEPA","WHO","USEPA","WHO","WHO","WHO","WHO")
mc_type <- c("Child Drinking Water Advisory",
             "Drinking Water Advisory",
             "Adult Drinking Water",
             "Recreational: Low Prob. of Effect",
             "Recreational: Moderate Prob. of Effect",
             "Recreational: High Prob. of Effect",
             "Recreational: Very High Prob. of Effect")

#wgt <- nla_dat$WGT_NLA
wgt<- rep(1,nrow(nla_dat))

@@

#mc_chla[1] <- get_chla(epa_child_cp,0.5)

@@

#mc_chla[2] <- get_chla(who_drink_cp,0.5)

@@

#mc_chla[3] <- get_chla(epa_adult_cp,0.5)

@@

#mc_chla[4] <- get_chla(who_rec_low1_cp,0.5)
epa_child <- NULL
who_drink <- NULL
epa_adult <- NULL
who_rec <- NULL
cps<- seq(0.1,0.9,0.1)
for(i in 1:9){
  epa_child[i] <- get_chla(epa_child_cp,cps[i])[1]
  who_drink[i] <- get_chla(who_drink_cp,cps[i])[1]
  epa_adult[i] <- get_chla(epa_adult_cp,cps[i])[1]
  who_rec[i] <- get_chla(who_rec_low1_cp,cps[i])[1]
}
mc_chla<-cbind(epa_child,who_drink,epa_adult,who_rec)

@@

```{r mc_chla_table, results='asis', echo=FALSE, tab.cap="Chlorophyll \\textit{a} concentrations that are associated with a 50% probability of exceeding a microcystin-LR health advisory concentration. \\label{tab:mc_chla_table}"}
mc_chla_tbl <- data.frame(seq(0.1,0.9,0.1),mc_chla)
names(mc_chla_tbl)<-c("Cond. Probability","USEPA Child", "WHO Drink", 
                      "USEPA Adult", "WHO Recreational")
#pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
knitr::kable(mc_chla_tbl)
```

### 2015-12-03 manuscript/manuscript.tex
closes #7 : creates table of chl a values across range of probability of exceeding and different advisory levels.

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper, we use this association to provide estimates
of chlorophyll \emph{a} concentrations that if exceeded would be
indicative of a higher probability of exceeding select health advisory
concentrations for microcystin-LR. Using the 2007 National Lakes
Assessment and a conditional probability approach, we identify
chlorophyll \emph{a} concentrations that are more likely than not to be
associated with an exceedance of a microcystin health advisory level. We
look at the recent US EPA standards for drinking water as well as the
World Health Organization levels for drinking water and recreational
use. For the specific advisory microcystin concentrations of 0.3, 1,
1.6, and 2, we find chlorophyll \emph{a} concentrations of 23.68, 63.94,
84.96, and 97.49, respectively. When managing for these various
microcystin levels, exceeding these reported chlorophyll \emph{a}
concentrations should be a trigger for further testing and possible
management action.

@@

One of these complications is that they rely on available measurements
of microcystin-LR. While laboratory testing remains the gold standard
for quantifying microcystin-LR concentrations in water samples, several
field test kits have been developed. Even though field tests provide a
much needed means for rapid assessment, they are not yet widely used and
are moderately expensive (approximately \$150-\$200 depending on
specific kit) with a limited shelf life (typically one year) (James et
al. 2011, Aranda-Rodriguez et al. 2015). Additionally, each technique
requires nuanced understanding of the detection method (e.g., limit of
detection, specific microcystin variants being measured, and sampling
protocol). Fortunately, microcystin-LR has been shown to be associated
with several other, more commonly measured and well understood
components of water quality.

Chlorophyll \emph{a} is a very commonly measured component of water
quality that is also known to be associated with Microsystin-LR
concentrations (Lee et al. 2000, Heisler et al. 2008, Paerl and Otten
2013). Additionally, there are many rapid measurements for assessing
chlorophyll \emph{a} levels \emph{in situ}. For instance, there are
small or hand held flourometers that measure chlorohpyll \emph{a}. Given
these facts, it might be possible to identify chlorophyll \emph{a}
concentrations that would be associated with the various Microcystin-LR
health advisory levels. Identifying these associations would provide
another tool for water resource managers to help manage the threat to
public health posed by cHABs and would be especially useful in the
absence of microcystin-LR concentrations. Thus, the goal of this paper
is to utilize the National Lakes Assessment (NLA) data and identify
chlorophyll \emph{a} concentrations that are associated with higher
probabilities of exceeding several microcystin-LR health advisory
concentrations (USEPA 2009). So that others may repeat or adjust this
analysis, the data, code, and this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

We used the 2007 NLA water quality and microcystin-LR concentration data
(USEPA 2009). These data represent a snapshot of water quality from the
summer of 2007 and data on chlorophyll \emph{a} and microcystin-LR
concentrations are available for lakes.

@@

We used a conditional probability analysis (CPA) approach to explore
associations between chlorophyll \emph{a} concentrations and World
Health Organization (WHO) and USEPA microcystin-LR health advisory
levels (Paul and Munns 2011). Many levels have been suggested (Table
\ref{tab:microcystin_levels}), but lakes with higher microcystin-LR
concentrations in the NLA were rare. Only 1.16 \% of lakes sampled had a
concentration greater than 10. Thus, for this analysis we focus on the
microcystin concentrations that are better represented in the NLA data.
These were 0.3, 1, 1.6, and 2 \(\mu\)g/L.

A detailed discussion of CPA is beyond the scope of this paper; see Paul
et al. (2005) and Hollister et al. (2008) for greater detail. For this
analysis, we used CPA to examine how the conditional probability of
exceeding one of the health advisory changes as chlorophyll \emph{a}
increases in a lake. The 95\% confidence intervals were calculated from
1000 bootstrapped samples. To identify chlorophyll \emph{a}
concentrations of concern we used a 50\% conditional probability of
exceeding each health advisory level and extracted the minimum
chlorophyll \emph{a} concentration that was associated with an upper
confidence level of 50\% or greater. As both microcystin-LR and
chlorophyll \emph{a} values were highely skewed right, a log base 10
transformation was used. Additional details of the specific
implementation are available at
\url{https://github.com/USEPA/microcystinchla}.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225.
Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent the detection
limits. Any value greater than that indicates the presence of
microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 3.17. Of all lakes sampled, 21\%
of lakes exceeded the USEPA childrens drinking water standard, 8.8\% of
lakes exceeded the USEPA adult drinking water standard,11.7\% of lakes
exceeded the WHO drinking water standard,and 7.3\% of lakes exceeded the
WHO recreational standard. For chlorophyll \emph{a}, the range was 0.07
to 936 \(\mu\)g/L. All lakes had reported chlorophyll \emph{a}
concentrations that exceeded detection limits. The median concentration
was 7.79 and the mean was 29.63. The associations between chlorophyll
\emph{a} and the upper confidence interval with a conditional
probability of 50\% were 23.68, 63.94, 84.96, and 97.49 for 0.3, 1.0,
1.6 and 2.0 \(\mu\)g/L advisory levels, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested
health advisory levels increases as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We use this association
to identify chlorophyll \emph{a} concentrations that are associated with
greater than even odds of exceeding a given health advisory level (Table
\ref{tab:mc_chla_table}). These represent 27.8\%, 11.9\%, 8.6\%, and
7.9\% of sample lakes for the U.S EPA childrens drinking water, the WHO
drinking water, the USEPA Adult drinking water, and the WHO recreational
advisories, respecitvely.

Furthermore, the chlorophyll \emph{a} cutoffs may be used to predict
whether or not a lake exceeds the microcystin-LR health advisories.
Doing so allows us to compare the accuracy of the prediction as well as
evaluate false negatives. Total accuracy of the four cutoffs predicting
microcystin-LR exceedances were 75\% for the USEPA children's advisory,
86\% for the WHO drinking water advisory, 90\% for the USEPA adult
advisory and, 90\% for the WHO regreational advisory (Tables
\ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table},
\ref{tab:adult_conmat_table}, \& \ref{tab:who_rec_conmat_table}).
However, total accuracy is only one part of the prediction performace
with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 9\%, 7\%, 6\% and , 5\% for the U.S EPA
childrens drinking water, the WHO drinking water, the USEPA adult
drinking water, and the WHO recreational advisories, respecitvely. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory.

There are numerous possible uses for the chlorophyll \emph{a} and
microcystin-LR advisory cut-off values. First, in the absence of
microcystin-LR measurements, exceedence of the chlorophyll \emph{a}
concentrations could be a trigger for further actions. Given that there
is uncertainity around these chlorophyll \emph{a} cutoffs the best case
scenario would be to monitor for chlorophyll and in the event of
exceeding a target concentration take water samples and have those
samples tested in a lab for microcystin-LR.

@@

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/epa_child_cp_plot-1.pdf}
\caption{Conditional probability plots showing association between the
probability of exceeding various microcystin-LR (MLR) health advisory
Levels. A.) Plot for USEPA childrens drinking water advisory (0.3
\(\mu\)g/L). B.) Plot for WHO drinking water advisory (1 \(\mu\)g/L).
C.) Plot for USEPA adult drinking water advisory (1.6 \(\mu\)g/L). D.)
Plot for WHO recreational advisory (2 \(\mu\)g/L).
\label{fig:multi_cp_plot}}
\end{figure}

@@

\begin{longtable}[c]{@{}llrr@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin-LR health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endfirsthead
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endhead
USEPA & Child Drinking Water Advisory & 0.3 & 23.68\tabularnewline
WHO & Drinking Water Advisory & 1.0 & 63.94\tabularnewline
USEPA & Adult Drinking Water & 1.6 & 84.96\tabularnewline
WHO & Recreational: Low Prob. of Effect & 2.0 & 97.49\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA
childrens drinking water advisory.
\label{tab:child_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 650 & 96\tabularnewline
TRUE & 161 & 121\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
drinking water advisory.
\label{tab:who_drink_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 835 & 75\tabularnewline
TRUE & 72 & 46\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA adult
drinking water advisory. \label{tab:adult_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 887 & 57\tabularnewline
TRUE & 50 & 34\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
recreational water advisory.
\label{tab:who_rec_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 903 & 48\tabularnewline
TRUE & 50 & 27\tabularnewline
\bottomrule
\end{longtable}

@@

Heisler, J., P. M. Glibert, J. M. Burkholder, D. M. Anderson, W.
Cochlan, W. C. Dennison, Q. Dortch, C. J. Gobler, C. A. Heil, E.
Humphries, and others. 2008. Eutrophication and harmful algal blooms: A
scientific consensus. Harmful algae 8:3--13.

@@

Lee, S., M.-H. Jang, H.-S. Kim, B.-D. Yoon, and H.-M. Oh. 2000.
Variation of microcystin content of microcystis aeruginosa relative to
medium n: P ratio and growth stage. Journal of Applied Microbiology
89:323--329.

@@

Paerl, H. W., and T. G. Otten. 2013. Harmful cyanobacterial blooms:
Causes, consequences, and controls. Microbial ecology 65:995--1010.

@@

@@

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper, we use this association to provide estimates
of chlorophyll \emph{a} concentrations that if exceeded would be
indicative of a higher probability of exceeding select health advisory
concentrations for microcystin-LR. Using the 2007 National Lakes
Assessment and a conditional probability approach, we identify
chlorophyll \emph{a} concentrations that are more likely than not to be
associated with an exceedance of a microcystin health advisory level. We
look at the recent US EPA standards for drinking water as well as the
World Health Organization levels for drinking water and recreational
use. For the specific advisory microcystin concentrations of 0.3, 1,
1.6, and 2, we find chlorophyll \emph{a} concentrations of 0.07, 0.07,
2.79, and 11.36, respectively. When managing for these various
microcystin levels, exceeding these reported chlorophyll \emph{a}
concentrations should be a trigger for further testing and possible
management action.

@@

One of these complications is that they rely on available measurements
of microcystin-LR. While laboratory testing remains the gold standard
for quantifying microcystin-LR concentrations in water samples, several
field test kits have been developed. Even though field tests provide a
much needed means for rapid assessment, they are not yet widely used and
are moderately expensive (approximately \$150-\$200 depending on
specific kit) with a limited shelf life (typically one year) (James et
al. 2011, Aranda-Rodriguez et al. 2015). Additionally, each technique
requires nuanced understanding of the detection method (e.g., limit of
detection, specific microcystin variants being measured, and sampling
protocol).

Fortunately, microcystin-LR has been shown to be associated with several
other, more commonly measured and well understood components of water
quality that are readily assessed in the field. For instance, there are
small or hand held fluorometers that measure chlorohpyll \emph{a}.
Additionally, chlorophyll \emph{a} is a very commonly measured component
of water quality that is also known to be positively associated with
Microsystin-LR concentrations (Pip and Bowman 2014, Yuan et al. 2014).
Yuan et. al (2014) explore these associations in detail and control for
other related variables. In their analysis they find that total nitrogen
and chlorophyll \emph{a} show the strongest association with
microcystin. Furthermore, they identify chlorophyll\emph{a} and total
nitrogen concentrations that are associated with exceeding 1 \(\mu\)g/L
of microcystin. Given these facts, it should be possible to identify
chlorophyll \emph{a} concentrations that would be associated with the
new USEPA Microcystin-LR health advisory levels for drinking water.
Identifying these associations would provide another tool for water
resource managers to help manage the threat to public health posed by
cHABs and would be especially useful in the absence of microcystin-LR
concentrations.

In this paper we build on past efforts and utilize the National Lakes
Assessment (NLA) data and identify chlorophyll \emph{a} concentrations
that are associated with higher probabilities of exceeding several
microcystin-LR health advisory concentrations (Chorus and Bartram 1999,
USEPA 2009, 2015). We add to past studies by exploring associations with
newly announced advisory levels and by also applying a different method,
conditional probability analysis. Utilizing different methods
strengthens the evidence for suggested chlorophyll \emph{a} levels that
are associated with increased risk of exceeding the health advisory
levels as those levels are not predicated on a single analytical method.
So that others may repeat or adjust this analysis, the data, code, and
this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

We used the 2007 NLA water quality and microcystin-LR concentration data
(USEPA 2009). These data represent a snapshot of water quality from the
summer of 2007 for the conterminous United States. Data on chlorophyll
\emph{a} and microcystin-LR concentrations are available for lakes.

@@

We used a conditional probability analysis (CPA) approach to explore
associations between chlorophyll \emph{a} concentrations and World
Health Organization (WHO) and USEPA microcystin-LR health advisory
levels (Paul and Munns 2011). Many levels have been suggested (Table
\ref{tab:microcystin_levels}), but lakes with higher microcystin-LR
concentrations in the NLA were rare. Only 1.16 \% of lakes sampled had a
concentration greater than 10. Thus, for this analysis we focus on the
microcystin concentrations that are better represented in the NLA data.
These were the USEPA children's drinking water advisory level of 0.3
\(\mu\)g/L (USEPA Child), the WHO drinking water advisory level of 1
\(\mu\)g/L (WHO Drinking), the USEPA adult drinking water advisory level
of 1.6 \(\mu\)g/L (USEPA Adult), and the WHO recreational, low
probability of effect advisory level of 2 \(\mu\)g/L (WHO Recreational).

Conditional probability analysis provides information about the
probability of observing one event given another event has also occured.
For this analysis, we used CPA to examine how the conditional
probability of exceeding one of the health advisory changes as
chlorophyll \emph{a} increases in a lake. We expect to find higher
chlorohpyll \emph{a} concentrations to be associated with higher
probabilities of exceeding the microcystin-LR health advisory levels. We
also caclulated bootstrapped 95\% confidence intervals (CI) using 1000
bootstrapped samples. Thus, to identify chlorophyll \emph{a}
concentrations of concern we identfiy the value of the upper 95\% CI
across a range of conditional probabilities of exceeding each health
advisory level. As both microcystin-LR and chlorophyll \emph{a} values
were highely skewed right, a log base 10 transformation was used.
Additional details of the specific implementation are available at
\url{https://github.com/USEPA/microcystinchla}. A more detailed
discussion of CPA is beyond the scope of this paper, but see Paul et al.
(2005) and Hollister et al. (2008) for greater detail.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225
\(\mu\)g/L. Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent
the detection limits. Any value greater than that indicates the presence
of microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 3.17. Of all lakes sampled, 21\%
of lakes exceeded the USEPA Child level, 8.8\% of lakes exceeded the
USEPA Adult level,11.7\% of lakes exceeded the WHO Drinking level,and
7.3\% of lakes exceeded the WHO Recreational level. For chlorophyll
\emph{a}, the range was 0.07 to 936 \(\mu\)g/L. All lakes had reported
chlorophyll \emph{a} concentrations that exceeded detection limits. The
median concentration was 7.79 \(\mu\)g/L and the mean was 29.63
\(\mu\)g/L. The associations between chlorophyll \emph{a} and the upper
confidence interval across a range of conditional probability values is
shown in Table \ref{tab:mc_chla_table}. Specific chlorophyll \emph{a}
that are associated with greater than even odds of exceeding the
advisory levels were 0.07, 0.07, 2.79, and 11.36 for 0.3, 1.0, 1.6 and
2.0 \(\mu\)g/L advisory levels, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested
health advisory levels increases as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We use this association
to identify chlorophyll \emph{a} concentrations that are associated with
greater than even odds of exceeding a given health advisory level (Table
\ref{tab:mc_chla_table}). These represent 99.9\%, 99.9\%, 77\%, and
43.8\% of sample lakes for the USEPA Child, the WHO Drinking, the USEPA
Adult, and the WHO recreational levels, respecitvely.

Furthermore, the chlorophyll \emph{a} cutoffs may be used to predict
whether or not a lake exceeds the microcystin-LR health advisories.
Doing so allows us to compare the accuracy of the prediction as well as
evaluate false negatives. Total accuracy of the four cutoffs predicting
microcystin-LR exceedances were 21\% for the USEPA children's advisory,
12\% for the WHO drinking water advisory, 31\% for the USEPA adult
advisory and, 61\% for the WHO regreational advisory (Tables
\ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table},
\ref{tab:adult_conmat_table}, \& \ref{tab:who_rec_conmat_table}).
However, total accuracy is only one part of the prediction performace
with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 0\%, 0\%, 0\% and , 1\% for the U.S EPA
childrens drinking water, the WHO drinking water, the USEPA adult
drinking water, and the WHO recreational advisories, respecitvely. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory.

There are numerous possible uses for the chlorophyll \emph{a} and
microcystin-LR advisory cut-off values. First, in the absence of
microcystin-LR measurements, exceedence of the chlorophyll \emph{a}
concentrations could be a trigger for further actions. Given that there
is uncertainity around these chlorophyll \emph{a} cutoffs the best case
scenario would be to monitor for chlorophyll \emph{a} and in the event
of exceeding a target concentration take water samples and have those
samples tested in a lab for microcystin-LR.

@@

\section{Acknowledgements}\label{acknowledgements}

@@

\begin{verbatim}
## Loading required package: grid
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/epa_child_cp_plot-1.pdf}
\caption{Conditional probability plots showing association between the
probability of exceeding various microcystin-LR (MLR) health advisory
Levels. A.) Plot for USEPA Child (0.3 \(\mu\)g/L). B.) Plot for WHO
Drinking (1 \(\mu\)g/L). C.) Plot for USEPA Adult (1.6 \(\mu\)g/L). D.)
Plot for WHO Recreational (2 \(\mu\)g/L). \label{fig:multi_cp_plot}}
\end{figure}

@@

\begin{longtable}[c]{@{}rrrrr@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin-LR health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Cond. Probability & USEPA Child & WHO Drink & USEPA Adult & WHO
Recreational\tabularnewline
\midrule
\endfirsthead
\toprule
Cond. Probability & USEPA Child & WHO Drink & USEPA Adult & WHO
Recreational\tabularnewline
\midrule
\endhead
0.1 & 0.07 & 0.07 & 0.07 & 1.47\tabularnewline
0.2 & 0.07 & 4.42 & 11.47 & 19.04\tabularnewline
0.3 & 2.79 & 16.32 & 30.62 & 52.70\tabularnewline
0.4 & 11.36 & 38.30 & 69.19 & 82.22\tabularnewline
0.5 & 23.68 & 65.20 & 84.96 & 103.90\tabularnewline
0.6 & 38.30 & 98.46 & 136.66 & 125.40\tabularnewline
0.7 & 68.85 & 133.20 & 871.20 & 871.20\tabularnewline
0.8 & 114.62 & 338.40 & 871.20 & 871.20\tabularnewline
0.9 & 198.72 & 516.00 & 871.20 & 871.20\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA
childrens drinking water advisory.
\label{tab:child_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 1 & 0\tabularnewline
TRUE & 810 & 217\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
drinking water advisory.
\label{tab:who_drink_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 1 & 0\tabularnewline
TRUE & 906 & 121\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA adult
drinking water advisory. \label{tab:adult_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 234 & 4\tabularnewline
TRUE & 703 & 87\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
recreational water advisory.
\label{tab:who_rec_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 566 & 15\tabularnewline
TRUE & 387 & 60\tabularnewline
\bottomrule
\end{longtable}

@@

@@

@@

@@

Pip, E., and L. Bowman. 2014. Microcystin and algal chlorophyll in
relation to nearshore nutrient concentrations in lake winnipeg, canada.
Environment and Pollution 3:p36.

@@

Yuan, L. L., A. I. Pollard, S. Pather, J. L. Oliver, and L. D'Anglada.
2014. Managing microcystin: Identifying national-scale thresholds for
total nitrogen and chlorophyll a. Freshwater Biology 59:1970--1981.
